---
title: "Final Project"
author: "Caleb Miller"
date: "2025-04-29"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Getting Started

First things first, I will have to load in tidyverse, download my data, and also get a look at my working directory

```{r}
#Load tidyverse

library(tidyverse)

#Getting my working directory

getwd()

#Reading in my data, it should already be placed in DataRaw, under XLSX

unzip(zipfile = "Data/DataRaw/XLSX/WI_Tree_Phenology.zip", exdir = "Data/DataRaw/XLSX/")
```

These are xlsx files, not csv files. Now, I could just open each file and save them as csv's, but I figured this is a good opportunity early-on to introduce a loop where I can take in the files individually and convert them to csv.

REVIEW: I'm not sure if this explanation is necessary; instead, you could just start with the readxl. According to Ellen, you could mention that this method won't work, so you have to do it the following way, but including all the code that didn't work isn't necessary.

```{r}
# install.packages("readxl") # This is the package needed to read the excel documents

#If you don't have the readxl package, you can download it by deleting the # in front of the install.packages("readxl") and running the code

```

### Starting the For Loop

```{r}
#Now we can get started with creating the for loop

library(readxl)

#I found a trick where I can assign directories names instead of typing every time

XLFolder <- "Data/DataRaw/XLSX/WI_Tree_Phenology/"
CSVFolder <- "Data/DataRaw/CSV/"

#Here you make the vector of the names of the files

XLFiles <- list.files(XLFolder, pattern = ".xlsx")
XLFiles

#Finally, the loop that creates the CSVs from XLSXs

for(i in XLFiles) { # Week13
  readxl::read_excel(paste0(XLFolder, i)) %>% #allows it to read in the XLSX doc
  write.csv(., paste0(CSVFolder, gsub(".xlsx", ".csv", i))) #tells it to take the XLSX doc, change it to a CSV by changing .xlsx to .csv, then put it in the CSV folder
}

```

Was that possibly more difficult than need be? Probably but I did it and I feel awesome.

Now, the next thing I like to do is get a good look at the docs. To do that, I'm going to read them in, but only get the heads of each one

```{r}
#Reading in the files themselves

SpringCoords <- read_csv("../Data/DataRaw/CSV/spring_coordinate_data.csv")
head(SpringCoords)
PhenoCodes <- read_csv("../Data/DataRaw/CSV/WI_Spring_Phenophase_Data_Codes.csv")
head(PhenoCodes)
TreeCoords <- read_csv("../Data/DataRaw/CSV/WI_Spring_Tree_Coordinates.csv")
head(TreeCoords)
PhenoAF <- read_csv("../Data/DataRaw/CSV/WI_Tree_Phenology_A_F.csv")
head(PhenoAF)
PhenoGL <- read_csv("../Data/DataRaw/CSV/WI_Tree_Phenology_G_L.csv")
head(PhenoGL)
PhenoMR <- read_csv("../Data/DataRaw/CSV/WI_Tree_Phenology_M_R.csv")
head(PhenoMR)
PhenoSX <- read_csv("../Data/DataRaw/CSV/WI_Tree_Phenology_S_X.csv")
head(PhenoSX)

PhenoSX

```

Great! Now we can look at all the data and come to some conculsions! Chiefly, there were certainly some decisions that were made with this data set. First and foremost, SpringCoords and TreeCoords are the same! Don't need two identical data tables in any situation. Further, all the Pheno data is broken up into 4 different tables sorted by A-F, G-L, M-R, and S-X. And on top of that, we do not need the 5th column at all and the data is all funky, specifically the way its writen includes the species of the tree, which is recorded in the TreeCoords (redundant data).

Other than that, there's other clean up (-99 instead of NA, excel messed up the date data) but first we need to format the data in a way that doesn't make my head hurt.

I believe the most important place to start would be merging all the Phenos into one table, that way whatever I do to it, I do to all of them.

Which is easier said than done, because these tables are Messed Up.

### Cleaning the Pheno Tables

```{r}
#First things first, I'm going to play around with PhenoAF a bit to see if i can format it the way I want

#I want to fix one thing right away: the really weird date columns and weird first column. But how do I do that?
#I don't need that first row, its just column names or data I have in other sheets
#Same with the Tree ID Column

CleanMyTables <- function(Table) { #Week 11
  RenameTable <- Table[-1, -5] %>% #Gets rid of unnecessary columns
    rename(Year = `...2`, #renames columns
         Date = `...3`,
         DOY = `...4`)
  RenameTable$DOY <- as.numeric(RenameTable$DOY) #changes DOY to numbers instead of characters
  CleanTable <- RenameTable %>%
    mutate(Date = as.Date(DOY-1, origin = paste0(Year, "-01-01"))) %>% #gets actual date using the DOY, replacing the messed-up Excel dates #Week 7
    select(!`...1`) #Gets rid of weird first column
  return(CleanTable)
}

PhenoAF <- CleanMyTables(PhenoAF)
PhenoGL <- CleanMyTables(PhenoGL) #Using my function to just clean the tables
PhenoMR <- CleanMyTables(PhenoMR)
PhenoSX <- CleanMyTables(PhenoSX)

PhenoSX # To check that the function worked

```

If you look at the data tables, each of the sites are named A1, A2, A3, B1, B2, B3, etc, and then the individual trees are named something like A1#1, A1#2, A1#3. Cleaning up these naming conventions will come next as it is not the best, but we will get to that. The most important to know right now is that PhenoAF has all the A through F sites, PhenoGL has all the G through L sites. The next thing we will do is merge them all into Pheno AX, which will contain all of the sites named A through X.

### Joining the Tables

Now: I still want to mix all the data together into one big table, so to do that, I'm going to use a full_join and join_by(Date)

```{r}
PhenoAX <- PhenoAF %>% #Week 4
  full_join(PhenoGL, join_by(Date, DOY, Year)) %>% #Organizes by Date, DOY, and Year. Otherwise, there will be duplicate DOY and Year columns (even though they all match by Date)
  full_join(PhenoMR, join_by(Date, DOY, Year)) %>% 
  full_join(PhenoSX, join_by(Date, DOY, Year))
PhenoAX
```

### Make Data Long, Not Wide

Now! This data is way too wide, and the columns are named after data (Tree and Site description). So now we are going to reassign them into a single column, named TreeID, and then we are going to split that column so that there is also a Site column (remember, A1#1 is a TreeID, with A1 denoting the site and #1 denoting the specific tree at that site).

```{r}
LongAX <- PhenoAX %>% #LongAX is the long version of PhenoAX #Week 6
  pivot_longer(`A1#1`:`X12#3`,
               names_to = "TreeID",
               values_to = "Phenophase") #The Pheno data needs to be added to a Phenophase column

head(LongAX)
tail(LongAX)
```

Now I am going to make a split from TreeID to add a Site column

```{r}
LongAX <- LongAX %>% 
  separate(TreeID, c("Site", "TreeID"), sep = "#")

head(LongAX)
tail(LongAX)
LongAX
```

### Correct NAs

Now that I have my one massive data sheet, I'm going to remove false NAs, recorded here as -99, and then I am going to drop any rows with NA in the Phenophase column. I am doing this because it is likely that certain sites were not monitored during a certain year, or maybe their leaves started opening later in the year, but this data is for observing the Phenophase, so if there is no data, I do not need that row.

```{r}
LongAX <- LongAX %>% 
  mutate(Phenophase = na_if(Phenophase, "-99")) %>% #changing -99s into NAs
  drop_na(Phenophase) #dropping the rows where Phenophase is NA

head(LongAX)
```

### Add Species Column

Now, to compare the phenology of the tree species against each other, I need to add a column for the species of each tree. To do taht, I am going to take the TreeCoords data, then split its TreeID column similar to what happened with LongAX, then I am going to join it to the dataframe as well

```{r}
TreeCoords <- TreeCoords %>% 
  separate(TreeID, c("Site", "TreeID"), sep = "#") %>%  #splitting Site and TreeID
  select(Site, TreeID, Species)

head(TreeCoords)

LongAX <- LongAX %>% 
  inner_join(TreeCoords, join_by(Site, TreeID))
head(LongAX)
```

Personally, I just would prefer to have species in front of Phenophase

```{r}
LongAX <- LongAX %>% 
  relocate(Species, .before = Phenophase)
head(LongAX)
```

### Sorting Phenophase

Now, Phenophase being 290 means nothing to me, so I'm going to assign it a quick explanation in a new column so that when I graph the data, anyone can read it and understand what it says. The problem with that is that there are many codes that are somewhat similar (Bud Swollen, 10-50% vs Bud Swollen, 50-90%) that I will likely push into a single category (Bud Swollen). Further, they are dissected into different categories dependent on whether they are deciduous or non-deciduous trees. So first thing I need to do is look at the species to know which are deciduous and not deciduous, then create an if-else function to sort the different phenophases into either their respective responses.

```{r}
LongAX %>%
  group_by(Species) %>% 
  summarize(abundance = n()) # Lets us see how many of each species there is, while also giving us a list of each species

#Create list of deciduous trees
Deciduous <- c("Aspen", "B. Aspen", "Basswood", "Blk. Ash", "R. Maple", "S. Alder", "S. Maple", "W. Birch", "Y. Birch")

#Final Dataframe with Phenology column describing the specific Phenology of the species

CompleteAX <- LongAX %>% #Week 12
  mutate(Phenology = case_when(substr(as.character(Phenophase), 1, 1) == "1" ~ "Bud Visible", #Makes Phenotype based on the first number of the Phenophase code
                               substr(as.character(Phenophase), 1, 1) == "2" ~ "Bud Swollen",
                               substr(as.character(Phenophase), 1, 1) == "3" ~ "Bud Open",
                               substr(as.character(Phenophase), 1, 1) == "4" & Species %in% Deciduous == TRUE ~ "Leaf Out", #Adds the condition of whether its deciduous or not
                               substr(as.character(Phenophase), 1, 1) == "4" & Species %in% Deciduous == FALSE ~ "Candle Out",
                               substr(as.character(Phenophase), 1, 1) == "5" & Species %in% Deciduous == TRUE ~ "Full Leaf Unfolded",
                               substr(as.character(Phenophase), 1, 1) == "5" & Species %in% Deciduous == FALSE ~ "Needles Unfolding from Candle",
                               substr(as.character(Phenophase), 1, 1) == "6" & Species %in% Deciduous == TRUE ~ "Leaf Expansion",
                               substr(as.character(Phenophase), 1, 1) == "6" & Species %in% Deciduous == FALSE ~ "Not Used")) #Only added the Not Used just in case there are some 600 stuff for non-deciduous trees.

CompleteAX
```

### Finally the Graphs

Finally I can graph the data, using DOY to show the season of leafing, filled with the particular Phenology, and then facet wrapped to show each species own phenology. I included a theme, as well as labelled my axes. 
```{r}
AXPlot <- ggplot(CompleteAX, mapping = aes(x = DOY, fill = Phenology)) + #Week 5
  geom_bar(alpha = 0.5) +
  labs(x = "Day of Year (DOY)", y = "Number of Trees", fill = "Phenology") +
  facet_wrap(~Species, scales = "free_y") +
  theme_classic()

AXPlot
```

From here, I can observe how many of each species were in each phenology class, and the progression of it throughout the season. Now, I am going to save it under in my Plots folder to complete my assignment. 

```{r}
ggsave("../Plots/AXPlot.png", AXPlot)
```

