---
title: "Final Project"
author: "Caleb Miller"
date: "2025-04-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Getting Started

First things first, I will have to load in tidyverse, download my data, and also get a look at my working directory
```{r}
#Load tidyverse

library(tidyverse)

#Getting my working directory

getwd()

#Reading in my data, it should already be placed in DataRaw, under XLSX

unzip(zipfile = "Data/DataRaw/XLSX/WI_Tree_Phenology.zip", exdir = "Data/DataRaw/XLSX/")

```
Uh oh, its already not working. Why? Because these are xlsx files, not csv files. Now, I could just open each file and save them as csv's, but I figured this is a good opportunity early-on to introduce a loop where I can take in the files individually and convert them to csv.

Is it probably simpler to just save the files? Yeah, probably.

Do I have to do extra work to get the proper packages? Yeah.

But I am justifying this as a nice way to get loops in here and expand my education a bit (I also just want to see if I can do it.)

```{r}
install.packages("readxl") # This is the package needed to read the excel documents
```

```{r}
#Now we can get started with creating the for loop
library(readxl)

#I found a trick where I can assign directories names instead of typing every time
XLFolder <- "Data/DataRaw/XLSX/WI_Tree_Phenology/"
CSVFolder <- "Data/DataRaw/CSV/"

#Here you make the vector of the names of the files
XLFiles <- list.files(XLFolder, pattern = ".xlsx")
XLFiles

#Finally, the loop that creates the CSVs from XLSXs
for(i in XLFiles) {
  readxl::read_excel(paste0(XLFolder, i)) %>% #allows it to read in the XLSX doc
  write.csv(., paste0(CSVFolder, gsub(".xlsx", ".csv", i))) #tells it to take the XLSX doc, change it to a CSV by changing .xlsx to .csv, then put it in the folder
}
```

